---
title: "EDA"
author: "Youlan Shen"
date: "2023-04-26"
output: pdf_document
---

# Data Exploratory Analysis

hurrican703.csv collected the track data of 703 hurricanes in the North Atlantic area since 1950. For all the storms, their location (longitude & latitude) and maximum wind speed were recorded every 6 hours. The data includes the following variables

1. **ID**: ID of the hurricans

2. **Season**: In which the hurricane occurred

3. **Month**: In which the hurricane occurred

4. **Nature**: Nature of the hurricane

* ET: Extra Tropical
* DS: Disturbance
* NR: Not Rated
* SS: Sub Tropical
* TS: Tropical Storm

5. **time**: dates and time of the record

6. **Latitude** and **Longitude**: The location of a hurricane check point

7. **Wind.kt** Maximum wind speed (in Knot) at each check point

```{r, message=FALSE, warning=FALSE}
# library all packages that we need at the beginning
library(tidyverse)
library(dplyr)
library(readxl)
library(car)
library(gtsummary)
library(corrplot)
library(caret)
```

## Summary table and Plot for hurricane data

```{r}
library(ggplot2)
dt = read.csv("hurrican703.csv")
ggplot(data=dt, aes(x = Longitude, y = Latitude)) + 
  stat_summary_2d(data = dt, aes(x = Longitude, y = Latitude, z = dt$Wind.kt), 
                  fun = median, binwidth = c(1, 1), show.legend = TRUE)
library(data.table)
dt <- as.data.table(dt)
summary(dt)
```

## Hurricane data on World Map

```{r}
library(maps)
map <- ggplot(data = dt, aes(x = Longitude, y = Latitude)) + 
  geom_polygon(data = map_data(map = 'world'), 
               aes(x = long, y = lat, group = group))
map +
  stat_summary_2d(data = dt, aes(x = Longitude, y = Latitude, z = dt$Wind.kt), 
                  fun = median, binwidth = c(1, 1), 
                  show.legend = TRUE, alpha = 0.75) + 
  ggtitle(paste0("Figure 1: Atlantic Windstorm mean knot"))
```

## Track of Each Hurricane on Map

```{r}
map <- ggplot(dt, aes(x = Longitude, y = Latitude, group = ID)) + 
  geom_polygon(data = map_data("world"), 
               aes(x = long, y = lat, group = group), 
               fill = "gray25", colour = "gray10", size = 0.2) + 
  geom_path(data = dt, aes(group = ID, colour = Wind.kt), size = 0.5) + 
  xlim(-138, -20) + ylim(3, 55) + 
  labs(x = "", y = "", colour = "Wind \n(knots)") + 
  theme(panel.background = element_rect(fill = "gray10", colour = "gray30"),
        axis.text.x = element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

seasonrange <- paste(range(dt[, Season]), collapse=" - ")

map + ggtitle(paste("Atlantic named Windstorm Trajectories (", 
                     seasonrange, ")\n")) 
```

## Track of Each Hurricane by Month on Map

```{r}
mapMonth <- map + facet_wrap(~ Month) +
  ggtitle(paste("Atlantic named Windstorm Trajectories by Month (", 
                 seasonrange, ")\n")) 
mapMonth
```

## Tidy Data and Preprocessing for MCMC

```{r}
# read in data from CSV file
hurricane <- read.csv("hurrican703.csv")
# tidy data on date
hurricane <- as_tibble(hurricane) %>%
  separate(time, into = c("Date", "Hour"), sep = " ") %>% 
  mutate(Hour = ifelse(Hour == "00:00:00)", 0, 
                       ifelse(Hour == "06:00:00)", 6,
                              ifelse(Hour == "12:00:00)", 12, 18))),
         Date = str_remove(Date, "\\("),
         Date = yday(Date),
         Month = factor(Month,levels = month.name))
# tidy data on latitude longitude wind_kt
hurricane <- hurricane %>% 
  group_by(ID) %>% 
  mutate(Lat_change =  Latitude - lag(Latitude, 1),
         Long_change = Longitude - lag(Longitude, 1),
         Wind_change = lag(Wind.kt, 1) - lag(Wind.kt, 2),
         Wind_prev = lag(Wind.kt, 1)) %>% 
  na.omit()
# save(hurricane, file = "hurricane.RData")
```

## Exploratory Plots of Hurricane in each 

```{r}
hurricane %>% 
  group_by(Month) %>% 
  summarise(count = n(),
            Ave.Speed = mean(Wind.kt)) %>% 
  ggplot(aes(x = Month)) +
  geom_col(aes(y = count), fill = "#69b3a2", alpha = 0.6) +
  geom_point(aes(y = Ave.Speed*100, color = " "), alpha = 0.6) +
  scale_y_continuous(
    name = "Frequncy of the Hurricane",
    sec.axis = sec_axis(~.*0.01, name = "Average Wind Speed")) +
  scale_color_manual(name = "Speed",
                     values = c(" " = "#404080")) +
  labs(title = "Figure 2: Hurricane Frequency and Average Wind Speed by Month")
```

```{r}
hurricane %>% 
  group_by(Season) %>% 
  summarise(count = n(),
            Ave.Speed = mean(Wind.kt)) %>% 
  ggplot(aes(x = Season)) +
  geom_col(aes(y = count), fill = "#69b3a2", alpha = 0.6) +
  geom_point(aes(y = Ave.Speed*10, color = " "), alpha = 0.6) +
  geom_smooth(aes(y = Ave.Speed*10, color = " "), alpha = 0.6) +
  scale_y_continuous(
    name = "Frequncy of the Hurricane",
    sec.axis = sec_axis(~.*0.1, name = "Average Wind Speed")) +
  scale_color_manual(name = "Speed",
                     values = c(" " = "#404080")) +
  labs(title = "Figure 3: Hurricane Frequency and Average Wind Speed by Year",
       x = "Year")
```

```{r}
hurricane %>% 
  group_by(Nature) %>% 
  ggplot(aes(x = Nature, y = Wind.kt)) +
  geom_boxplot(fill = "#69b3a2", alpha = 0.6) +
  labs(title = "Figure 4: Hurricane Wind Speed Boxplot by Nature",
       x = "Nature")
```




